Constructing ordinal outcomes for clinical trials

### 2025-06-06 constructing_ordinal_outcomes.tex
Follman et al 2020 new outcome approach

  \item \textbf{Hierarchical composition} Complete tie-breaking of $A$ by
    $B$ (``prioritized outcomes''; lexicographic ordering) within
    \emph{every} level of $A$.

    This is what win methods do.

@@

These approaches implicitly \textbf{bundle together three different things}:
\begin{enumerate}
  \item Constructing an implicit ordinal outcome through
    hierarchical composition (complete tie-breaking/prioritized outcomes)
  \item
    Targeting a nonparametric effect measure defined via the
    ``pairs'' device (win
    probability/win ratio/win odds/etc.)
  \item Estimating that effect measure in a
    nonparametric/flexible manner using pairs, often also with devices for
    accounting for censoring
\end{enumerate}

@@

@@

\begin{itemize}
  \item Situating methods in their proper context can help
    trialists make informed choices among them
  \item Win methods can make use of
    plots and diagnostics for general ordinal outcomes, for the \emph{implicit}
    ordinal variable
    \begin{itemize}
      \item comparative CDF or ``survival''
        (complementary CDF) plots for assessing stochastic
        monotonicity \todo{this
          seems not done routinely? perhaps because of the difficulty
        of incorporating censoring?}
      \item ridit plots
        \citep{brossHowUseRidit1958,jansenRiditAnalysisReview1984} relative to a
        reference group, which should allow one to display both full
        eCDF information
        and also win statistics (based on the interpretation of the
          mean ridit as a win
        probability)
    \end{itemize}e.g.
  \item Ordinal regression models can use win
    methods for model checking; it is also possible to present a model-based win
    effect measure alongside a common odds ratio for example. See
    \citep{agrestiOrdinalProbabilityEffect2017} for example, which covers both
    latent variable and predicted probability versions.
  \item The outcomes
    \emph{implicitly} constructed using DOOR and win methods may contain many
    levels, and may contain levels where an effect is not so meaningful (e.g.
    time-to-death when the followup time is short). By \emph{explicitly}
    constructing the associated ordinal outcome one might be forced to be more
    aware of this
  \item What is a reasonable win analysis of a specific component
    score? From the traditional ordinal endpoint school of thought
    what would make
    most sense is to \emph{only tie-break for the lowest and highest
      levels of that
    score}, and e.g. plot eCDFs of this new variable. Is this what the plots do
    that break down wins and losses by component score?
\end{itemize}

>>>>

    This is what win methods do \todo{I \emph{THINK}. Need to go
      through an example very carefully to check it's not in fact
      sequential composition. Check
      \citep{follmannAnalysisOrderedComposite2020} and the $x$ axis of
    their survival plots}
  \item \textbf{Hierarchical composition} Complete tie-breaking of $A$ by
    $B$ (``prioritized outcomes''; lexicographic ordering) within
    \emph{every} level of $A$.

@@

These approaches implicitly \textbf{bundle together three different things}:
\begin{enumerate}
  \item Constructing an implicit ordinal outcome through
    hierarchical composition (complete tie-breaking/prioritized outcomes)
  \item
    Targeting a nonparametric effect measure defined via the
    ``pairs'' device (win
    probability/win ratio/win odds/etc.)
  \item Estimating that effect measure in a
    nonparametric/flexible manner using pairs, \textbf{with devices for
    accounting for censoring}
\end{enumerate}

@@

\citet{follmannAnalysisOrderedComposite2020} is an example of the
approach of explicitly constructing a new ordinal outcome based on a
hierarchy (in this case a hierarchy of 3 times-to-event, with event
times for worse events overriding event times for better ones.).

@@

\begin{itemize}
  \item Situating methods in their proper context can help
    trialists make informed choices among them
  \item Win methods can make use of
    plots and diagnostics for general ordinal outcomes, for the \emph{implicit}
    ordinal variable
    \begin{itemize}
      \item comparative CDF or ``survival''
        (complementary CDF) plots for assessing stochastic
        monotonicity. \citet{follmannAnalysisOrderedComposite2020}
        also does this.
      \item ridit plots
        \citep{brossHowUseRidit1958,jansenRiditAnalysisReview1984} relative to a
        reference group, which should allow one to display both full
        eCDF information
        and also win statistics (based on the interpretation of the
          mean ridit as a win
        probability)
    \end{itemize}e.g.
  \item Ordinal regression models can use win
    methods for model checking; it is also possible to present a model-based win
    effect measure alongside a common odds ratio for example. See
    \citep{agrestiOrdinalProbabilityEffect2017} for example, which covers both
    latent variable and predicted probability versions.
  \item The outcomes
    \emph{implicitly} constructed using DOOR and win methods may contain many
    levels, and may contain levels where an effect is not so meaningful (e.g.
    time-to-death when the followup time is short). By \emph{explicitly}
    constructing the associated ordinal outcome one might be forced to be more
    aware of this
  \item What is a reasonable win analysis of a specific component
    score? From the traditional ordinal endpoint school of thought
    what would make
    most sense is to \emph{only tie-break for the lowest and highest
      levels of that
    score}, and e.g. plot eCDFs of this new variable. Is this what the plots do
    that break down wins and losses by component score?
\end{itemize}
